---
title: "Howto Use and Extend the `MsBackendRawFileReader`?"
author:
- name: Tobias Kockmann
  affiliation:
    - &id Functional Genomics Center Zurich - Swiss Federal Institute of Technology in Zurich
  email: Tobias.Kockmann@fgcz.ethz.ch
- name: Christian Panse
  affiliation:
    - *id
    - Swiss Institute of Bioinformatics
  email: cp@fgcz.ethz.ch
ppackage: MsBackendRawFileReader
output:
  BiocStyle::html_document
abstract: |
  `r BiocStyle::Githubpkg("fgcz/MsBackendRawFileReader")` implements an
  MsBackend for the `r BiocStyle::Biocpkg("Spectra")` package using
  Thermo Fisher Scientific's NewRawFileReader .Net libraries.
  The package is generalizing the functionallity provided by the
  `r BiocStyle::Biocpkg("rawrr")`. The vignette utilizes data provided through
  the `r BiocStyle::Biocpkg("tartare")` package.
vignette: |
  %\VignetteIndexEntry{Howto use and extend the package?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography:
  - MsBackendRawFileReader.bib
---

# Requirements

```{r require}
stopifnot(require(Spectra),
          require(MsBackendRawFileReader),
          require(tartare),
          require(BiocParallel))
```

# Load data

```{r tartareEH4547, warning=FALSE, message=FALSE, eval=TRUE}
# fetch via ExperimentHub
library(ExperimentHub)
eh <- ExperimentHub::ExperimentHub()
```

```{r tartare}
query(eh, c('tartare'))
```

```{r EH3220, message=FALSE, warning=FALSE, echo=FALSE}
EH3220 <- normalizePath(eh[["EH3220"]])
(rawfileEH3220 <- paste0(EH3220, ".raw"))
if (!file.exists(rawfileEH3220)){
  file.link(EH3220, rawfileEH3220)
}

EH3222 <- normalizePath(eh[["EH3222"]])
(rawfileEH3222 <- paste0(EH3222, ".raw"))
if (!file.exists(rawfileEH3222)){
  file.link(EH3222, rawfileEH3222)
}

EH4547  <- normalizePath(eh[["EH4547"]])
(rawfileEH4547  <- paste0(EH4547 , ".raw"))
if (!file.exists(rawfileEH4547 )){
  file.link(EH4547 , rawfileEH4547 )
}
```

```{r EH3219}
EH3219 <- normalizePath(eh[["EH3219"]])
EH3221 <- normalizePath(eh[["EH3221"]])
EH4547 <- normalizePath(eh[["EH4547"]])
```

# On using the `MsBackendRawFileReader`

call the constructor

```{r backendInitialize, message=FALSE}
beRaw <- Spectra::backendInitialize(
  MsBackendRawFileReader::MsBackendRawFileReader(),
  files = c(rawfileEH3220, rawfileEH3222, rawfileEH4547))
```

```{r show}
beRaw
```


# Application example

Here we reproduce the Figure 2 of @rawrr `r BiocStyle::Biocpkg("rawrr")`.
The `r BiocStyle::Githubpkg("fgcz/MsBackendRawFileReader")` ships with a 
`filterScan` method  using functionality provided by the C# libraries by 
Thermo Fisher Scientific @rawfilereader.


```{r filterTest, eval=FALSE}
SF1 <- (beRaw |>  
   filterScan("ms2 487.2567"))

SF2 <- beRaw |>  
   filterScan("ms2 487.25") |>
   sparceVector() |> 
   dotProduct(list(mZ=c(1, 2, 3), intensity=c(10, 100, 100)))

# output 

# dotProduct and index vom backend

```

```{r rawrrFigure2, fig.retina=3}
(S <- (beRaw |>  
   filterScan("FTMS + c NSI Full ms2 487.2567@hcd27.00 [100.0000-1015.0000]") )[437]) |> 
  plotSpectra()

# supposed to be scanIndex 9594
S

# add yIonSeries to the plot
(yIonSeries <- protViz::fragmentIon("LGGNEQVTR")[[1]]$y[1:8])
names(yIonSeries) <- paste0("y", seq(1, length(yIonSeries)))
abline(v = yIonSeries, col='#DDDDDD88', lwd=5)
axis(3, yIonSeries, names(yIonSeries))
```

# Extend Class

For demonstration reasons, we extent the `MsBackend` class by a simple filter method.
`filterIons` returns spectra iff all fragment ions, given as argument, match. We use `r BiocStyle::CRANpkg("protViz")``::findNN` binary search
method for determining the nearest mZ peak for each ion. If the mass error between an ion and an mz value is less than the given mass tolerance, an ion is considered a hit.
 
```{r defineFilterIon, echo=FALSE}
setGeneric("filterIons", function(object, ...) standardGeneric("filterIons"))

setMethod("filterIons", "MsBackend",
          function(object, mZ=numeric(), tol=numeric(), ...) {
            
            keep <- lapply(peaksData(object, BPPARAM = bpparam()),
                           FUN=function(x){
                             NN <- protViz::findNN(mZ, x[, 1])
                             hit <- (error <- mZ - x[NN, 1]) < tol & x[NN, 2] >= quantile(x[, 2], .9)
                             #print(hit)
                             if (sum(hit) == length(mZ))
                               TRUE
                             else
                               FALSE
                           })
            
            object[unlist(keep)]
          })
```

apply the defined `filterIons` method

```{r filterIons2, fig.retina=3, fig.height=12, echo=FALSE}
beRaw |> 
  filterScan("FTMS + c NSI Full ms2 487.2567@hcd27.00 [100.0000-1015.0000]") |>
  filterIons(yIonSeries, tol = 0.0005) |> 
  plotSpectra()
```


# Benchmark on determine break-even of split size

```{r benchmark, eval=FALSE, echo=FALSE}
b <- lapply(c(1000,5000,10000), function(n){
  lapply(c(8,16,32,64,128,256,8,16,32,64,128,256,8,16,32,64,128,256,8,16,32,64,128,256,8,16,32,64,128,256,8,16,32,64,128,256), function(size){
    start_time <- Sys.time()
    rv <- MsBackendRawFileReader:::.RawFileReader_read_peaks(rawfileEH4547, 1:n, maxGroupSize = size)
    end_time <- Sys.time()
    d <- end_time - start_time
    message(sprintf("%d - %s", size, d))
    data.frame(size=size, time=d/n, n=n, workers=bpparam()$workers)
  })
})

```


```{r readSpecs}
fn <- file.path(system.file(package = 'MsBackendRawFileReader'),
               'extdata', 'specs.csv')
specs.readSpectra <- read.csv2(fn, header=TRUE)
```


```{r lattice::xyplot, echo=FALSE, fig.retina=3}
lattice::xyplot((as.numeric(time)) * workers~ size | factor(n) ,
                group= host,
                data=specs.readSpectra,
                horizontal=F,
                auto.key = TRUE,
                layout=c(3,1),
                ylab='time [seconds] to read a spectrum',
                xlab='number of spectra / file')
```

# Session information

```{r si}
sessionInfo()
```

# References
