---
title: "Howto use the MsBackendRawFileReader package?"
author:
- name: Christian Panse
  affiliation: Functional Genomics Center Zurich
  email: cp@fgcz.ethz.ch
- name: Tobias Kockmann
  affiliation: Functional Genomics Center Zurich
  email: tobias.Kockmann@fgcz.ethz.ch
package: MsBackendRawFileReader
output:
  BiocStyle::html_document
abstract: |
  xxx xxx xxx `r Githubpkg("rformassspectrometry/Spectra")`  [@Huber2015, @Trachsel2018]
vignette: |
  %\VignetteIndexEntry{Howto use the package?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography:
  - MsBackendRawFileReader.bib
---

# Usage

## Read a raw file

setup

```{r librarySnippet, message=FALSE}
library(MsBackendRawFileReader)

library(knitr)

(fls <- file.path(path.package(package = 'MsBackendRawFileReader'),
    'extdata', 'sample.raw'))
```

initialize and populate the backend

```{r HelloWorldSnippet, message=FALSE}
be <- backendInitialize(MsBackendRawFileReader(), files = fls, extra=TRUE)
```

the output of the show method:

```{r Spectra}
(S <- Spectra(be))
```

## `spectraVariables` 

```{r spectraVariables}
spectraVariables(S)
```

## `spectraData`
```{r sample}

df <- spectraData(S)
kable((df[2,]),  longtable = TRUE)
```

for those who want to have fun please continue reading.


```{r peaks}
peaks(S)
```

## Some statistics

```{r msLevel}
table(msLevel(S))
```


## Some graphics

### precursorMz count

```{r histogram, fig.retina=3}
library(lattice)
histogram(~precursorMz(S) | precursorCharge(S) * basename(dataStorage(S)),
          type = 'count')
```


### LC-MS map
```{r lcmsmap, fig.retina=3}
plot(rtime(S), precursorMz(S),
     main = basename(dataOrigin(S)[1]),
     pch = 16,
     col = rainbow(5, alpha = 0.5)[1+precursorCharge(S)])
```



# Misc

## Availability of `spectraVariables` 

The following table provides an overview of the from `spectraVariables` returned
columns using differen `MsBackend` classes.

```{r compareSpectraVariables, echo=FALSE, warning=FALSE, message=FALSE}
library(mzR)
fls.mzR <- file.path(path.package(package = 'MsBackendRawFileReader'),
    'extdata', 'sample.mzXML')
be.mzR <- backendInitialize(MsBackendMzR(), files = fls.mzR, extra=TRUE)


sv.rfr <- spectraVariables(S)
sv.mzR <- data.frame(spectraVariables=spectraVariables(Spectra(be.mzR)))
sv.mzR$Attr <- TRUE

sv.rfr <- data.frame(spectraVariables=spectraVariables(S))
sv.rfr$Attr <- TRUE
M <-merge(sv.rfr, sv.mzR, by='spectraVariables', all.x=T, all.y=T)
names(M) <- c('spectra Variables', 'RawFileReader', 'mzR')
kable(M)
```


## Diagnostic plots

to the benefit of diagnostic plot functions [@Trachsel2018] use the following
method to coerce the Spectra object into an `r Githubpkg("fgcz/rawDiag")`
object.
Off note, the plots have been designed mainly for Orbitrap devices but are not
limited to it.

```{r rawDiag, eval=FALSE}
RAW <- as.rawDiag(S)
```

## Generate a mascot generic file (mgf)

```{r mgf, eval=FALSE}
mgf(S, filename = 'sample.mgf')
```


## Generate a `peaklistSet`

to apply the `r CRANpkg("protViz")` functions.

```{r protViz, eval=FALSE}
PLS <- as.peaklistSet(S) 
```


## Testing multiple files

```{r testingMultipleFiles}
rv <- lapply(1:2, function(x){
  file.copy(from = fls, to = tempfile(fileext='.raw'))})


be <- backendInitialize(MsBackendRawFileReader(),
  files = file.path(tempdir(), list.files(path = tempdir(), pattern = 'raw$')))

table(basename(dataOrigin(be)))

#  the rDotNet package tested only with one worker.
register(SnowParam(workers = 1, type = "SOCK") , default = TRUE); registered()
SS <- Spectra(be)
```

```{r TicBasePeakversusRtime, fig.retina=3}
plot(tic(S) ~ rtime(S), type='l')
lines(basePeakIntensity(S)~rtime(S), col='red')
```


```{r cycleNumbers}
plot(S@backend@rawfileReaderObj[[1]]$GetCycleNumbers() ~  rtime(S), type='l')
```

# Session Info

```{r sessionInfo}
sessionInfo()
```

#References


