---
title: "Howto use the MsBackendRawFileReader package?"
author:
- name: Christian Panse
  affiliation: Functional Genomics Center Zurich
  email: cp@fgcz.ethz.ch
- name: Tobias Kockmann
  affiliation: Functional Genomics Center Zurich
  email: tobias.Kockmann@fgcz.ethz.ch
package: MsBackendRawFileReader
output:
  BiocStyle::html_document
abstract: |
  xxx xxx xxx `r Githubpkg("rformassspectrometry/Spectra")`  [@Huber2015, @Trachsel2018]
vignette: |
  %\VignetteIndexEntry{Howto use the package?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography:
  - MsBackendRawFileReader.bib
---

# Usage

## Read a raw file

setup
```{r librarySnippet, message=FALSE}
library(MsBackendRawFileReader)
library(knitr)
(fls <- file.path(path.package(package = 'MsBackendRawFileReader'),
                                'extdata', 'sample.raw'))
```

initialize and populate the backend
```{r HelloWorldSnippet, message=FALSE}
be <- backendInitialize(MsBackendRawFileReader(), files = fls)
```

the output of the show method:
```{r Spectra}
(S <- Spectra(be))
```


```{r spectraVariables}
spectraVariables(S)
```


```{r sample}
df <- spectraData(S)
kable(cbind(df[2,]@rownames, df[2,]@listData))
```

for those who want to have fun please continue reading.


```{r peaks}
peaks(S)
```

## Some statistics

```{r msLevel}
table(msLevel(S))
```


## Some graphics

### precursorMz count

```{r histogram, fig.retina=3}
library(lattice)
histogram(~precursorMz(S) | precursorCharge(S) * basename(dataStorage(S)),
          type = 'count')
```


### LC-MS map
```{r lcmsmap, fig.retina=3}
plot(rtime(S), precursorMz(S),
     main = basename(dataOrigin(S)[1]),
     pch = 16,
     col = rainbow(5, alpha = 0.5)[1+precursorCharge(S)])
```



# Misc


## Diagnostic plots

to the benefit of diagnostic plot functions [@Trachsel2018] use the following
method to coerce the Spectra object into an `r Githubpkg("fgcz/rawDiag")`
object.
Off note, the plots have been designed mainly for Orbitrap devices but are not
limited to it.

```{r rawDiag, eval=FALSE}
RAW <- as.rawDiag(S)
```

## Generate a mascot generic file (mgf)

```{r mgf, eval=FALSE}
mgf(S, filename = 'sample.mgf')
```


## Generate a `peaklistSet`

to apply the `r CRANpkg("protViz")` functions.

```{r protViz, eval=FALSE}
PLS <- as.peaklistSet(S) 
```

## Debugging



### manually stating the `CLRServer.exe`
```{bash eval=FALSE}
/Library/Frameworks/Mono.framework/Commands/mono64 \
  --llvm /Library/Frameworks/R.framework/Versions/3.6/Resources/library/rDotNet/server/bin/Debug/CLRServer.exe \
  -url svc://localhost:56789/ \
  -dll /Library/Frameworks/R.framework/Versions/3.6/Resources/library/MsBackendRawFileReader/exec/MsBackendRawFileReader.dll
```


### reading basic data structs
```{r debugging, eval=TRUE}
library(MsBackendRawFileReader)

(rawfile <- file.path(path.package(package = 'MsBackendRawFileReader'),
                                'extdata', 'sample.raw'))
```


call the constructor for a given rawfile
```{r debuggingCallConstructor}
x <- .cnew ("Rawfile", rawfile)
```

call methods of the C# class
```{r debuggingHaveFun}
x$GetMsLevel(1)

vapply(1:10, FUN=function(z){x$GetMsLevel(z)}, FUN.VALUE=as.integer(1))

head(x$GetSpectrumMz(1, ""))

```

### Extra Header Information

```{r debuggingExtraHeaderInformation}
dfExtraHeader <- DataFrame(label=x$GetTrailerExtraHeaderInformationLabel(),
                           scan1=x$GetTrailerExtraHeaderInformationValue(1),
                           scan2=x$GetTrailerExtraHeaderInformationValue(2),
                           scan100=x$GetTrailerExtraHeaderInformationValue(100))

```

```{r kableExtraHeader, echo=FALSE, comment='GetTrailerExtraHeaderInformation'}
kable(dfExtraHeader)
```

### On using filters

see also [@RawfileReader2015, page 28]

```{r filters}
x$GetAutoFilters()

x$IsValidFilter("most likely not a valid filter string")
x$IsValidFilter("ms2 776.9297@hcd28.00")
```

## Testing multiple files

```{r testingMultipleFiles}
rv <- lapply(1:2, function(x){
  file.copy(from = fls, to = tempfile(fileext='.raw'))})


be <- backendInitialize(MsBackendRawFileReader(),
  files = file.path(tempdir(), list.files(path = tempdir(), pattern = 'raw$')))

table(basename(dataOrigin(be)))

#  the rDotNet package tested only with one worker.
register(SnowParam(workers = 1, type = "SOCK") , default = TRUE); registered()
SS <- Spectra(be)
```

# Session Info

```{r sessionInfo}
sessionInfo()
```

#References


