---
title: "Howto use the MsBackendRawFileReader package?"
author:
- name: Christian Panse
  affiliation: Functional Genomics Center Zurich
  email: cp@fgcz.ethz.ch
- name: Tobias Kockmann
  affiliation: Functional Genomics Center Zurich
  email: tobias.Kockmann@fgcz.ethz.ch
package: MsBackendRawFileReader
output:
  BiocStyle::html_document
abstract: |
  xxx xxx xxx `r Githubpkg("rformassspectrometry/Spectra")`  [@Huber2015, @Trachsel2018]
vignette: |
  %\VignetteIndexEntry{Howto read Thermo-Fisher raw files?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography:
  - MsBackendRawFileReader.bib
---

# Usage

## Read a raw file

setup
```{r librarySnippet, message=FALSE}
library(MsBackendRawFileReader)

(fls <- file.path(path.package(package = 'MsBackendRawFileReader'),
                                'extdata', 'sample.raw'))

```

initialize and populate the backend
```{r HelloWorldSnippet, message=FALSE}
be <- backendInitialize(MsBackendRawFileReader(), files = fls)
```

the output of the show method:
```{r}
(S <- Spectra(be))
```

for those who want to have fun please continue reading.

## Some statistics

```{r}
table(msLevel(S))
```

## Some graphics


### LC-MS map
```{r lcmsmap, fig.retina=3}
plot(rtime(S), precursorMz(S),
     main = basename(dataOrigin(S)[1]),
     pch = 16,
     col = rainbow(5, alpha = 0.5)[1+precursorCharge(S)])
```



# Misc


## Diagnostic plots

to the benefit of diagnostic plot functions [@Trachsel2018] use the following
method to coerce the Spectra object into an `r Githubpkg("fgcz/rawDiag")`
object.
Off note, the plots have been designed mainly for Orbitrap devices but are not
limited to it.

```{r rawDiag, eval=FALSE}
RAW <- as.rawDiag(S)
```

## Generate a mascot generic file (mgf)

```{r mgf, eval=FALSE}
mgf(S, filename = 'sample.mgf')
```


## Generate a `peaklistSet`

to apply the `r CRANpkg("protViz")` functions.

```{r protViz, eval=FALSE}
PLS <- as.peaklistSet(S) 
```

## Debugging

```{r eval=FALSE}
library(MsBackendRawFileReader)

(rawfile <- file.path(path.package(package = 'MsBackendRawFileReader'),
                                'extdata', 'sample.raw')))

x <- .cnew ("Rawfile", rawfile)

x$GetMsLevel(1)

vapply(1:10, FUN=function(z){x$GetMsLevel(z)}, FUN.VALUE=as.integer(1))

x$GetSpectrumMz(1, "")
```

# Session Info

```{r}
sessionInfo()
```

#References


